Machine Learning Methods in Functional Programming and Interactive Theorem Proving

### 2015-10-20 TransferReport/report.tex
More report

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{amssymb}
%\usepackage{enumitem}
\usepackage{paralist}
\usepackage{csquotes}
\usepackage[affil-it]{authblk}

@@

The automation of ``routine'' mathematical reasoning by computers is increasingly important for both mathematics and computer science. What has previously been seen as a pure research field of academic interest TODO: better wording [CITE] is becoming an indispensable tool at the cutting edge of these fields. The controversy surrounding Hales' TODO proof of the Kepler conjecture, and its subsequent formalisation [CITE], is perhaps the most famous example of mathematical results becoming too large and intricate to be feasibly verified, let alone discovered, without computer assistance.

At the same time, the modern increasing reliance of a proof which relies on computers to such an extent that it is in cannot be understood  requirement
 - Maths, verification, etc. are important
 - Theorem proving: ATP and ITP

Any theorem proving approach is fundamentally limited by its reliance on the user to provide the conjectures to be (dis)proved. We focus This

 - (Automated) theory exploration
 - Theory of ``interestingness'' is missing

TODO: Include work trying to assign probability to formal sentences

In \S\ref{background}, we formalise the notions of \emph{theory}, \emph{theory exploration} and \emph{interestingness}, along with neccessary background such as \emph{automated theorem proving}. In \S\ref{related} we review related work, including the application of statistical methods to formal systems and the use of pure exploration in artificial intelligence domains. \S\ref{current} summarises our recent work undertaken in this area, and \S\ref{future} presents future directions this can be taken.

@@

@@

Before formalising To frame the concepts we will be building on, we first present an informal, simplistic picture. followed by individual define our terms in turn. Take a typical Gentzen-style judgement, e.g. in type theory:

@@

TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to

TODO: Background section

TODO: Reword this, theory exploration is our main point. It is a research direction/area, etc. Don't define it negatively (``without goals''), define it positively. Maybe stop trying to encompass everyone's different demarcations, and have a ``for our purposes, ....''

When computers are used to aid the development of formal proofs, it is usually within either the \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP) framework. These approaches are characterised as being \emph{goal-driven}: particular conjectures must be provided by the user before attempts are made to (dis)prove them. In this work we focus on the automation of the conjecture generation process itself, a task we refer to as \emph{theory exploration}. We can characterise theory

This reliance on goals In addition to these traditional approaches, we also consider the related recently proposed \emph{theory exploration} (TE) paradigm, which does not use (explicit) goals, and of which ATP and ITP are crucial components.

\subsection{Interactive Theorem Proving}

The simplest approach to proof automation automated ITP systems are built around an automated \emph{proof checker}. This is a relatively simple task to mechanise \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic (for example, TODO: Reference some example, eg. proof of false in Coq. A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  TODO Maybe Chapter 18. Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers). instead?. This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.

In the ITP setting, proving a conjecture is equivalent to producing an input to the checker which \begin{inparaenum}[a)]
  \item is accepted and
  \item has that conjecture as its conclusion.
\end{inparaenum}

TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?

Of course, \emph{producing} such proofs is much more difficult than checking them; as the name suggests, ITP leaves this task up to the user. To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format. This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, there are practical benefits to turning theorem proving into programming. Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process. Secondly, the process of \emph{software verification}, a common task for formal proof, can be achieved in a very straightforward way: the software can be implemented inside the proof language, certified correct by the proof checker, then \emph{extracted} into an executable form.

Despite these conveniences, ITP still requires enormous effort for non-trivial proof or verification tasks (e.g. see \cite{hales2015formal}). The most obvious way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible. The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which translates goals into a form suitable for a multitude of ATP systems and invokes them all in parallel to see if any can provide a proof which Isabelle's core checker will accept.

\subsection{Automated Theorem Proving}

@@

TODO: Related work section
\section{Related Work}
\label{related}

@@

\subsubsection{Artificial Curiosity}

@@

@@

@@

\subsubsection{Concept Formation}

Alan Bundy et al

Eurisko, AM, etc.?

Genetic programming? GP in finite algebras

@@

\section{CURRENT PROGRESS}
\label{current}

Divide and conquer for theory exploration
Does it help?

Clustering and feature extraction

\section{FUTURE WORK}
\label{future}

@@

How about systems based on term rewriting, logic programming, etc.?

@@

\section{The \textsc{ML4HS} Framework}\label{ml4hs}

@@

Thank you to the \textsc{HipSpec} team at Chalmers University (Moa Johansson,
Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful
discussions of these ideas.
\fi

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{amssymb}
%\usepackage{enumitem}
\usepackage{paralist}
\usepackage{csquotes}
\usepackage[affil-it]{authblk}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,decorations.pathreplacing,calc}

@@

As computers and software become more capable, and as our reliance on them increases, the importance of \emph{understanding}, \emph{predicting} and \emph{verifying} these systems grows; which is undermined by their ever-increasing complexity. Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}. By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis. However, the question still remains: on what should we focus that analysis?

In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems, especially software, as a way to make productive use of resources in an often intractable domain. Appeals to interestingness arise when more direct measures, such as utility, are not available; as an extreme example, we could this is analogous to for example, we could characterise the distinction between pure and applied mathematics. much work in pure mathematics could be considered For example, the inclusion of particular statements in a program or proof development can be easily justified by whether they contribute to the solution; however, in a \emph{library} there is no particular problem to be solved, in which case statements may instead be judged on how interesting they are. Appeals to interestingness abound in the history of computer-assisted reasoning; for example Lenat's 1977 description of the AM system using information to \textquote{concentrate upon the most interesting [concepts]} \cite{lenat1977automated}. Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.

We begin in \S \ref{background} by defining a formal framework for our investigation, and how it relates to the existing theorem proving landscape. In \S \ref{conceptformation} we consider those heuristics proposed in the literature to date, whilst \S \ref{related} casts a wider net to encompass pure exploration and interestingness in other domains of machine learning and artificial intelligence. We consider a selection of results from the theorem proving community in \S \ref{examples}, both to evaluate these heuristics and to identify common features. We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.

@@

We will focus our investigation on inductive theorems regarding programs in the Haskell language, since this is an area with existing work \cite{Claessen_hipspec:automating} against which we can compare our methods.

\begin{figure}
  \centering
  \tikzstyle{decision} = [diamond, draw, fill=blue!20, text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
  \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text centered, rounded corners, minimum height=4em]
  \tikzstyle{line} = [draw, -latex']
  \tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm, minimum height=2em]

  \begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
    \node [block] (cf) {Concept Formation};
    \node [block, below of=cf] (cg) {Conjecture Generation};
    \node [block, below of=cg] (tp) {Theorem Proving};

    % Draw edges
    \path [line] (cf) -- (cg);
    \path [line] (cg) -- (tp);

    % Draw braces
    \draw [decorate,decoration={brace,amplitude=10pt, mirror}]
    (cf.north west) -- (tp.south west) node [black,midway,xshift=-1.5cm]
    {\footnotesize ITP};

    \draw [decorate,decoration={brace,amplitude=10pt}]
    (tp.north east) -- (tp.south east) node [black,midway,xshift=0.3cm]
    {\footnotesize ATP};

    \draw [decorate,decoration={brace,amplitude=10pt}, xshift=5cm]
    ($ (cg.north east) + (1,0) $) -- ($ (tp.south east) + (1,0) $)
    node [black,midway, xshift=0.6cm] {\footnotesize Theory Exploration};
  \end{tikzpicture}
  \caption{Core stages of computer-assisted proof \label{proofprocess}}
\end{figure}

@@

Before defining formalising To frame the concepts we will be building on, we first present an informal, simplistic picture. followed by individual define our terms in turn. Take a typical Gentzen-style judgement, e.g. in type theory:

@@

\fi

% TODO: Maybe focus on one language, system, set of algorithms, etc.? That way, we can make things concrete, rather than ``what about edge-case XYZ?''. HipSpec calls out to

% TODO: Background section

% TODO: Reword this, theory exploration is our main point. It is a research direction/area, etc. Don't define it negatively (``without goals''), define it positively. Maybe stop trying to encompass everyone's different demarcations, and have a ``for our purposes, ....''

To set the scene for the formal definitions which follow, we first give an overview of our key terms and concepts, which are also depicted in figure \ref{proofprocess}. Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP). Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characteris both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures). In this work, we push the focus back to the prior steps of \emph{concept formation} (\S \ref{conceptformation}) and \emph{conjecture generation} (\S \ref{conjecturegeneration}), which are characterised as being \emph{open-ended} and \emph{creative}, in that solutions must be judged (as interesting or not) rather than merely constructed.

\subsection{Interactive Theorem Proving}
\label{itp}

The ITP process relies on a \emph{proof checker} $C$, which is a decision procedure for determining if a given proof object $P$ witnesses a given statement $S$:

$$ C : (P \times S) \rightarrow Boolean $$

The ITP process then consists of finding an appropriate $P$ for the given \emph{goal} $S$, which is accepted by $C$:

$$ ITP : (S \times C) \rightarrow \{ P | C(P, S) = True \} $$

Automating a proof checker is relatively simple \cite{boyer1994qed}, although care must be taken during implementation since bugs may cause inconsistencies to arise in the resulting logic \iffalse (for example, TODO: Reference some example, eg. proof of false in Coq). \fi A popular approach to mitigating this risk, as well as making meta-level analysis of the system more tractable, is known as the \emph{de Bruijn criterion}, which means the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}  \iffalse TODO Maybe Chapter 18. Proof-Assistants Using Dependent Type Systems (Henk Barendregt, Herman Geuvers). instead?. \fi This is the approach taken by, for example, the Isabelle and Coq systems \cite{nipkow2002isabelle} \cite{bertot2013interactive}.

% TODO: How closely does Isabelle match the type-theoretic approach of Coq, Agda, etc.? Can we make this less ``fluffy''?
%      Maybe ``for our purposes'' focus on type-based ITP, untyped ATP, etc.?

Of course, \emph{producing} such proofs is much more difficult than checking them; as the name suggests, ITP leaves this task up to the user. To facilitate easier proving, many ITP systems such as Coq and Agda use a pure functional programming language as their proof format. This is theoretically elegant due to the \emph{Curry-Howard correspondence} \cite{wadler2015propositions}, and there are also practical benefits to turning theorem proving into programming. Firstly, many techniques developed in software engineering such as meta-programming (usually in the form of \emph{tactics}), encapsulation and modularity can be directly applied to the proving process. Secondly, the process of \emph{software verification}, a common task for formal proof, can be achieved in a very straightforward way: the software can be implemented inside the proof language, certified correct by the proof checker, then \emph{extracted} into an executable form.

Despite these conveniences, ITP still requires enormous effort for non-trivial proof or verification tasks (e.g. see \cite{hales2015formal}). One common way to mitigate this problem is by implementing powerful tactics to automate as much of a proof as possible \iffalse Adam Chlipala \fi. The most striking example of this approach is the \emph{Sledgehammer} system in Isabelle/HOL \cite{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.

\subsection{Automated Theorem Proving}
\label{itp}

@@

\section{Examples of Interesting Objects}
\label{examples}

Before exploring abstract definitions of interestingness, we can first consider some examples of theorems, lemmas and properties which occur in existing developments, to see what real users considered interesting enough to formalise. This will give an empirical basis for our intuitions, which we can compare with the definitions to follow.

\subsection{Generalisations}

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of which $S$ is a special case. Proofs by induction often require generalisation, since the specific obligations which arise during the proof may be unsuitable for the inductive hypotheses which are available. However, we cannot blindly generalise all obligations we encounter, since \emph{over-generalising} results in obligations which are too strong to prove at all. We must therefore rely on heuristics to choose when and how to generalise, which can in part consider how interesting the generalisations are.

As an example, we can take Boyer and Moore's proof of the associativity of multiplication in ACL2 \cite{boyer1983proof}:

$$(x * y) * z = x * (y * z)$$

During the course of the proof, the following obligation arises:

\begin{equation}
  \tag{conc3}
  (y + (x * y)) * z = (y * z) + ((x * y) * z)
  \label{eq:conc3}
\end{equation}

ACL2 automatically generalises \eqref{eq:conc3} by replacing the repeated sub-term $x * y$ with a fresh variable $w$:

\begin{equation}
  \tag{conc4}
  (y + w) * z = (y * z) + (w * z)
  \label{eq:conc4}
\end{equation}

This generalised form is clearly the distributivity law for multiplication and addition, which can be proved separately to the original goal of associativity.

\subsection{Auxilliary Lemmas}

Many of the above generalisations, such as  do not rely on context

\subsection{QuickCheck Properties}

\section{Related Work}
\label{related}

@@

\subsubsection{Artificial Curiosity}
\label{curiosity}

@@

The core problem of assigning ``interestingness'' to logical formulae is the application of statistical reasoning to the discrete, semantically-rich domain of formal systems. This problem has been tackled from various directions for a variety of reasons; here we summarise those contributions which seem of particular importance for theory exploration.

@@

The combinatorial nature of formal systems leads many algorithms in this domain, such as proof search methods, to have exponential time complexityCITE; hence even a modest size increase can turn a trivial problem into an intractable one. It is currently infeasible to improve the scaling behaviour of many such algorithms; for example, problems like TODO are in the infamous NP-complete complexity classCITE. On the other hand, we can turn this difficulty around: a modest \emph{decrease} in size may turn an intractable problem into a solvable one. We can ensure that the solutions to these reduced problems coincide with the original if we only remove \emph{redundant} information. This leads to the idea of \emph{relevance filtering}.

Relevance filtering simplifies a proof search problem by removing from consideration those axioms, definitions, lemmas, etc. which are deemed \emph{irrelevant}. The technique is used in Sledgehammer during its translation of Isabelle/HOL theories to statements in first order logic: rather than translating the entire theory, only those lemmas which are deemed ``relevant'' (required for proving the statement) are included. This reduces the size of the problem and speeds up the proof search, but it creates the new problem of determining when a lemma is relevant: how do we know what will be required, before we have the proof?

Sledgehammer's approach is to use \emph{supervised learning} to approximate the relevance relation: a naive bayes classifier is trained on TODO

@@

\subsubsection{Interestingness in Concept Formation}
\label{conceptformation}

% Alan Bundy et al

% Eurisko, AM, etc.?

% Genetic programming? GP in finite algebras

@@

\section{Contributions}
\label{current}

\subsection{The \textsc{ML4HS} Framework}\label{ml4hs}

% Divide and conquer for theory exploration
% Does it help?

% Clustering and feature extraction

\section{Future Work}
\label{future}

@@

How about systems based on term rewriting, logic programming, etc.?
\fi

\section{Conclusion}
\label{conclusion}

@@

@@

Thank you to the \textsc{HipSpec} team at Chalmers University (Moa Johansson,
Koen Claessen, Nick Smallbone, Dan Ros{\'e}n and Irene Lobo Valbuena) for useful
discussions of these ideas.

\fi
